---
title: "cyclistic_bike_share_Q3"
author: "Emily Jelagat"
output: html_document
date: "2024-01-30"
---

# Cyclistic_Bike_Share_Q3_Analysis.

This analysis is a case study from Google Data Analytics Certificate course. The datasets were made available by Motivate International Inc.
In this project, I will be merging datasets from July, August and September 2020 to get a single dataset for Q3 2020, conduct simple analysis to help answer the key question "In what ways do member and casual riders use cyclistic bikes differently?"


I cleaned and performed some calculations of the datasets on Microsoft Excel, before importing them to R i.e
 - I dropped Columns representing the ‘starting latitude’, ‘ending latitude’ ‘starting longitude’ ‘ending longitude’ of rides were removed as they were not necessary for the business task.
 - Records of data with negative ride_length were removed because they were taken out of the docks for quality checks.
 - I added 3 rows to help solve the business task i.e ‘ride_length’, ‘member_casual’, ‘day_of_week’.


# Install required packages.

tidyverse is for data import and wrangling 

```{r}
install.packages("tidyverse")
```

```{r}
library("tidyverse")
```

readr is used for reading rectangular data e.g csv file

```{r}
install.packages("readr")
```
```{r}
library("readr")
```

ggplot is for visualizations
```{r}
install.packages("ggplot2")
```
```{r}
library(ggplot2) 
```


libridate for date functions
```{r}
install.packages("libridate")
```
```{r}
library("libridate")
```

dplyr help complete some data manipulation tasks
```{r}
install.packages("dplyr")
```
```{r}
library("dplyr")
```

works with functions and vectors to help with coding
```{r}
install.packages("purrr")
library("purrr")
```

# Select working directory with Q3 csv files
 
I selected my working directory using the session tab.

getwd() will wisplay my working directory
```{r}
getwd()
```
# 1. COLLECT DATA 

## Import csv cyclistic bike share q3 data to global environment
 
 July 2020 dataset
 
```{r}
library(readr)
cyclistic_bike_share_july_2020 <- read_csv("cyclistic_bike_share_july_2020.csv")
View(cyclistic_bike_share_july_2020)
```


August 2020 dataset

```{r}
library(readr)
cyclistic_bike_share_august_2020 <- read_csv("cyclistic_bike_share_august_2020.csv")
View(cyclistic_bike_share_august_2020)
```

September 2020 dataset

```{r}
library(readr)
cyclistic_bike_share_sept_2020 <- read_csv("cyclistic_bike_share_sept_2020.csv")
View(cyclistic_bike_share_sept_2020)
```




# 2. WRANGLE DATA AND COMBINE INTO A SINGLE FILE

## Inspected and changed data type on the files tab

I converted ride_id and rideable_type to character in all datasets so that they can stack correctly.


Stacked cyclistic data from July, August, September 2020 to get Q3 data.
I used bind_rows function to merge the data from July,August, September 2020 to get Q3 dataframe.
       
```{r}
cyclistic_bike_share_Q3_2020 <- bind_rows(cyclistic_bike_share_july_2020,cyclistic_bike_share_august_2020,cyclistic_bike_share_sept_2020)
View(cyclistic_bike_share_Q3_2020)
```


# 3. CLEANING Q3 DATA

I installed the 'skimr' and 'janitor' packages for data cleaning.
skimr helps to summarize data

```{r}
install.packages("skimr")
library("skimr")
```

janitor helps in data cleaning

```{r}
install.packages("janitor")
library("janitor")
```


## Inspect new Q3 table
```{r}
colnames(cyclistic_bike_share_Q3_2020)
```

## number of rows in Q3 
```{r}
nrow(cyclistic_bike_share_Q3_2020)
```

## dimensions of the q3 dataframe
```{r}
dim(cyclistic_bike_share_Q3_2020)
```

## see list of columns and data types in q3 dataset
```{r}
str(cyclistic_bike_share_Q3_2020)
```


## Summary of Q3 data

```{r}
summary(cyclistic_bike_share_Q3_2020)
```

# Covert ride_length to numeric to run claculations on the data

```{r}
is.factor(cyclistic_bike_share_Q3_2020$ride_length)
```

```{r}
cyclistic_bike_share_Q3_2020$ride_length <-as.numeric(as.character(cyclistic_bike_share_Q3_2020$ride_length))
```

To confirm if the ride_length column has changed to numeric run this code

```{r}
is.numeric(cyclistic_bike_share_Q3_2020$ride_length)
```

```{r}
cyclistic_bike_share_Q3_2020_2 <-cyclistic_bike_share_Q3_2020_complete = na.omit(cyclistic_bike_share_Q3_2020$ride_length)
```

## remove ride length with NA values and create new data frame

Install the tidyr package to help remove NA values in the ride_length column

```{r}
install.packages("tidyr")
library("tidyr")
```

```{r}
cyclistic_bike_share_Q3_2020_2 <- cyclistic_bike_share_Q3_2020 %>% drop_na(ride_length)
```


## arrange q3 data in ascending order based on the time the ride started.

```{r}
library("dplyr")
```

```{r}
cyclistic_bike_share_Q3_2020_2 <- arrange(cyclistic_bike_share_Q3_2020_2, started_at)
```


# 4. CONDUCT DESCRIPTIVE ANALYSIS

## mean ride_length (all figures in seconds)

```{r}
mean(cyclistic_bike_share_Q3_2020_2$ride_length)
```
**1587.013**

## average ride length in seconds
```{r}
median(cyclistic_bike_share_Q3_2020_2$ride_length)
```
**953**

## longest ride in seconds 
```{r}
max(cyclistic_bike_share_Q3_2020_2$ride_length)
```
**86389**

## Shortest ride in seconds ignoring the missing values

```{r}
min(cyclistic_bike_share_Q3_2020_2$ride_length)
```
**0**

##compare member and casual riders

```{r}
aggregate(cyclistic_bike_share_Q3_2020_2$ride_length ~ cyclistic_bike_share_Q3_2020_2$member_casual, FUN =mean)
```

```{r}
aggregate(cyclistic_bike_share_Q3_2020_2$ride_length ~ cyclistic_bike_share_Q3_2020_2$member_casual, FUN = median)
```
```{r}
aggregate(cyclistic_bike_share_Q3_2020_2$ride_length ~ cyclistic_bike_share_Q3_2020_2$member_casual, FUN = max)
```
```{r}
aggregate(cyclistic_bike_share_Q3_2020_2$ride_length ~ cyclistic_bike_share_Q3_2020_2$member_casual, FUN = max)
```

##Average ride time for members and casual riders by day of the week

```{r}
aggregate(cyclistic_bike_share_Q3_2020_2$ride_length ~ cyclistic_bike_share_Q3_2020_2$member_casual+cyclistic_bike_share_Q3_2020_2$day_of_week, FUN = mean)
```



#Export q3 dataset to desktop folder csv files for further analysis and visualizations in Tableau 

```{r}
write.csv(cyclistic_bike_share_Q3_2020_2, "C:\\Users\\emily\\Desktop\\csv files\\cyclistic_bike_share_Q3_2020.csv", row.names = FALSE)
```



